---
title: "Hotgenes API and Methods"
output:
  github_document:
    html_preview: false
vignette: >
  %\VignetteIndexEntry{Hotgenes API and Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

This vignette covers the core API for interacting with a Hotgenes object
after it has been created. All examples use the pre-built `dds_Hotgenes`
object that ships with the package (created from a DESeq2 analysis of
Ewing sarcoma RNAseq data).

For instructions on creating a Hotgenes object from scratch, see
**01 Creating Hotgenes Objects**.

```{r set, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  fig.path  = "02_API_and_Methods_files/figure-",
  fig.height = 4, fig.width = 6
)
```

```{r load, message=FALSE}
library(Hotgenes)

# Load the pre-built DESeq2-based example
htgs <- readRDS(
  system.file("extdata", "dds_Hotgenes.RDS",
              package = "Hotgenes",
              mustWork = TRUE)
) |> update_object()
```

---

## 1. Object Summary

Calling a Hotgenes object prints a summary of its contents:

```{r show}
htgs
```

---

## 2. Accessing Slots

### Sample metadata

```{r coldata}
coldata_(htgs)
```

### Normalized expression data

`Normalized_Data_()` returns the named list of expression matrices.
Use `ExpressionSlots_()` to see available slot names.

```{r expression_slots}
ExpressionSlots_(htgs)
```

```{r expression_data}
# Access a specific slot by name
Normalized_Data_(htgs, slot = "vsd")[1:4, 1:4]
```

### Sample IDs

```{r sample_ids}
SampleIDs_(htgs)
```

### Feature alias mapper

```{r mapper}
Mapper_(htgs) |> head()
```

### Design matrix

```{r design_matrix}
designMatrix_(htgs) |> head()
```

### Original DE object

`O_()` returns the original analysis object stored at creation time
(a `DESeqDataSet` for DESeq2-derived objects, an `EList` with an
embedded limma fit for limma-derived objects).

```{r original_object}
O_(htgs) |> class()
```

### Auxiliary assays

Auxiliary assays store per-sample data that do not fit into the
expression matrix—clinical scores, QC metrics, low-throughput assay
values, etc. The slot starts empty (only a `SampleIDs` column).

```{r aux_view}
auxiliary_assays_(htgs)
```

Add data with the assignment form of the accessor:

```{r aux_add}
set.seed(42)
n_samples <- length(SampleIDs_(htgs))

new_aux <- auxiliary_assays_default(htgs) |>
  dplyr::mutate(
    score_A = rnorm(n_samples, mean = 5, sd = 1),
    score_B = runif(n_samples, min = 0, max = 10)
  )

auxiliary_assays_(htgs) <- new_aux
auxiliary_assays_(htgs)
```

---

## 3. The `DE()` Function

`DE()` is the main entry point for querying differential expression
results. It supports filtering by contrast, adjusted p-value,
log2 fold-change, and gene name.

### Default: all significant features across all contrasts

```{r de_details}
DE(htgs, padj_cut = 0.05, .log2FoldChange = 1) |> head()
```

### Select specific contrasts

```{r de_contrasts}
# List available contrasts
contrasts_(htgs)
```

```{r de_one_contrast}
DE(htgs,
   contrasts      = "sh_EWS_vs_Ctrl",
   padj_cut       = 0.05,
   .log2FoldChange = 1) 
```

### Look up specific genes with `hotList`

When `hotList` is provided, all matches are returned regardless of
significance thresholds. A `significant` column is added to flag
features that pass the thresholds.

```{r de_hotlist}
DE(htgs,
   hotList        = c("CSF2", "IL6", "CCL2"),
   padj_cut       = 0.05,
   .log2FoldChange = 0.5)
```

### `Report` options

The `Report` argument controls what `DE()` returns:

| `Report` | Returns |
|---|---|
| `"Details"` (default) | Complete DE result table |
| `"Features"` | Character vector of feature names per contrast |
| `"contrast_dir"` | Feature names named by `<contrast>_up` / `<contrast>_down` |
| `"Length"` | Number of significant features per contrast |
| `"Ranks"` | Named numeric vector of stats (for GSEA input) |
| `"FC"` | Named numeric vector of log2 fold changes |

```{r de_length}
DE(htgs, Report = "Length", padj_cut = 0.1)
```

```{r de_features}
DE(htgs,
   Report   = "Features",
   contrasts = "sh_EWS_vs_Ctrl",
   # limit to top 5
   Topn = 5,
   padj_cut = 0.1) 
```

```{r de_ranks}
# Ranks are sorted by the stat column (used as GSEA input)
DE(htgs,
   Report    = "Ranks",
   contrasts  = "sh_EWS_vs_Ctrl",
   
   # limit to top 5
   Topn = 5,
   Rank_name = "Feature",
   padj_cut  = 1) 
```

```{r de_fc}
DE(htgs,
   Report    = "FC",
   contrasts  = "sh_EWS_vs_Ctrl",
   # limit to top 5
   Topn = 5,
   Rank_name = "Feature",
   padj_cut  = 1) 
```

---

## 4. `Output_DE_()` — Low-level Access

`Output_DE_()` returns the raw list of DE tables stored in the object,
without further filtering or formatting. Useful when you need to pass
results downstream in list form.

```{r output_de}
raw_DE <- Output_DE_(htgs, padj_cut = 1, as_list = TRUE)
names(raw_DE)
raw_DE[["sh_EWS_vs_Ctrl"]] |> head(3)
```

---

## 5. `DExps()` — Expression Data Merged with Metadata

`DExps()` joins normalized expression values with sample metadata,
making it easy to build plots or run downstream models without
extra data-wrangling steps.

```{r dexps}
DExps(htgs,
      hotList    = c("IL6", "CSF2"),
      coldata_ids = c("Hrs", "sh"),
      Query_set  = TRUE) |> head()
```

---

## 6. `DECoefs()` — Per-Feature Coefficients

`DECoefs()` extracts the model coefficients (one column per contrast)
for a set of features of interest.

```{r decoefs}
DECoefs(htgs, hotList = c("CSF2", "IL6", "CCL2"))
```

---

## 7. Updating the Mapper

The mapper table can be replaced or augmented at any time:

```{r mapper_update}
current_mapper <- Mapper_(htgs)
head(current_mapper)

# Add a synthetic annotation column
updated_mapper <- current_mapper |>
  dplyr::mutate(custom_note = paste0("gene:", .data$Feature))

Mapper_(htgs) <- updated_mapper
Mapper_(htgs) |> head()
```

---

## 8. Combining Multiple Hotgenes Objects

When you have objects from different experiments or platforms, store
them in a named list. The Shiny app and several reporting functions
accept this list directly.

```{r combine}
# Load the limma-based example as a second object
fit_Hotgenes <- readRDS(
  system.file("extdata", "fit_Hotgenes.RDS",
              package = "Hotgenes",
              mustWork = TRUE)
)

Hotgenes_list <- list(
  DESeq2_experiment = htgs,
  limma_experiment  = fit_Hotgenes
)

# Access individual objects from the list
Hotgenes_list[["limma_experiment"]]
```

---

## Summary of Key Accessors

| Function | Purpose |
|---|---|
| `coldata_()` | Sample metadata data.frame |
| `Normalized_Data_()` | Named list of expression matrices |
| `ExpressionSlots_()` | Names of available expression slots |
| `SampleIDs_()` | Sample identifiers |
| `Features_()` | Feature (gene/protein) names |
| `contrasts_()` | Available contrast names |
| `Mapper_()` | Feature alias mapping table |
| `designMatrix_()` | Model design matrix |
| `contrastMatrix_()` | Contrast matrix |
| `O_()` | Original DE object |
| `auxiliary_assays_()` | Auxiliary sample-level data |
| `DE()` | Query DE results with flexible reporting |
| `Output_DE_()` | Raw DE list (low-level) |
| `DExps()` | Expression + metadata joined table |
| `DECoefs()` | Per-feature model coefficients |
