---
title: "Visualization and Exploration"
output:
  github_document:
    html_preview: false
vignette: >
  %\VignetteIndexEntry{Visualization and Exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

This vignette demonstrates the visualization functions available in
Hotgenes. All examples use `fit_Hotgenes`, the pre-built limma-based
example object that ships with the package.

For details on creating Hotgenes objects, see **01 Creating Hotgenes
Objects**. For details on the API, see **02 API and Methods**.

```{r set, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  fig.path  = "03_Visualization_and_Exploration_files/figure-",
  fig.height = 4, fig.width = 6
)
```

```{r load, message=FALSE}
library(Hotgenes)

fit_Hotgenes <- readRDS(
  system.file("extdata", "fit_Hotgenes.RDS",
              package = "Hotgenes",
              mustWork = TRUE)
)
```

---

## 1. `DEPlot()` — Overview of All Contrasts

`DEPlot()` gives a bird's-eye view of DE results across all contrasts
in the object. Each bar shows the number of significant features for
one contrast, split by direction (up / down).

```{r dep_1}
DEPlot(fit_Hotgenes, .log2FoldChange = 0, padj_cut = 0.1)
```

Pass a `hotList` to highlight specific features across contrasts:

```{r dep_2}
DEPlot(fit_Hotgenes,
       hotList        = c("CSF1", "IL6"),
       .log2FoldChange = 0,
       padj_cut       = 0.1)
```

---

## 2. `VPlot()` — Volcano Plots

`VPlot()` renders a standard volcano plot for a single contrast. Points
are coloured by significance and fold-change. Labels are added for the
top hits (or for features in `hotList`).

```{r vplot_1}
VPlot(fit_Hotgenes,
      contrasts       = "sh_EWS_vs_Ctrl",
      .log2FoldChange = 1,
      padj_cut        = 0.1)
```

Highlight a gene of interest with `hotList`:

```{r vplot_2}
VPlot(fit_Hotgenes,
      contrasts       = "sh_EWS_vs_Ctrl",
      .log2FoldChange = 1,
      padj_cut        = 0.1,
      hotList         = "CSF2",
      Hide_labels     = FALSE)
```

---

## 3. `Venn_Report()` — Overlapping Hits Across Contrasts

`Venn_Report()` identifies features that are significant in more than
one contrast (or contrast direction). It returns both a Venn diagram
and the underlying intersection tables.

Use `Report = "Features"` to overlap by feature name (ignoring
direction), or `Report = "contrast_dir"` to treat up- and down-regulated
hits as separate sets (maximum two contrasts).

```{r venn_features, fig.height=6, fig.width=6}
# Overlapping features (ignoring direction) across three contrasts
venn_out <- fit_Hotgenes |>
  DE(
    Report   = "Features",
    contrasts = c("sh_EWS_vs_Ctrl", "Hrs_2_vs_0", "Hrs_6_vs_0"),
    padj_cut = 0.1
  ) |>
  Venn_Report(set_name_size = 4, stroke_size = 0.5, text_size = 4)

venn_out$vennD
```

```{r venn_dir, fig.height=6, fig.width=6}
# Including directionality (up/down) for two contrasts
venn_dir_out <- fit_Hotgenes |>
  DE(
    Report   = "contrast_dir",
    contrasts = c("sh_EWS_vs_Ctrl", "Hrs_6_vs_0"),
    padj_cut = 0.1
  ) |>
  Venn_Report(set_name_size = 3.5)

venn_dir_out$vennD
```

Retrieve the names and gene lists from the intersections:

```{r venn_results}
# Names of features found in all intersecting sets
venn_out$Names

# All intersection sets as a named list
venn_out$Intsect |> head()
```

---

## 4. `DEphe()` — Heatmap of Top Hits

`DEphe()` generates a pheatmap of the top `Topn` features for a
selected contrast, annotated with sample metadata.

```{r heatmap_1, fig.height=8, fig.width=8}
DEphe(fit_Hotgenes,
      contrasts         = "sh_EWS_vs_Ctrl",
      Topn              = 5,
      cellheight        = 10,
      cellwidth         = 8,
      annotation_colors = coldata_palettes(fit_Hotgenes),
      annotations       = c("Hrs", "sh"))
```

Use `label_by` to replace the default Feature IDs with any alias
column in the mapper:

```{r heatmap_2, fig.height=8, fig.width=8}
DEphe(fit_Hotgenes,
      contrasts         = "sh_EWS_vs_Ctrl",
      label_by          = "ensembl_id",
      Topn              = 5,
      cellheight        = 10,
      cellwidth         = 8,
      annotation_colors = coldata_palettes(fit_Hotgenes),
      annotations       = c("Hrs", "sh"))
```

Subset samples on the fly with `SampleIDs`:

```{r heatmap_3, fig.height=6, fig.width=6}
selected_samples <- SampleIDs_(fit_Hotgenes)[1:8]

DEphe(fit_Hotgenes,
      contrasts         = "sh_EWS_vs_Ctrl",
      Topn              = 5,
      SampleIDs         = selected_samples,
      cellheight        = 10,
      cellwidth         = 8,
      annotation_colors = coldata_palettes(fit_Hotgenes),
      arrangeby         = c("Hrs", "sh"),
      annotations       = c("Hrs", "sh"))
```

---

## 5. `ExpsPlot()` — Individual Gene Expression Plots

`ExpsPlot()` plots the expression trajectory for one or more features
across samples, coloured and faceted by metadata variables. Expression
data and coldata are joined automatically.

```{r exps_1}
ExpsPlot(fit_Hotgenes,
         xVar    = "Hrs",
         yVar    = c("CSF2", "IL6"),
         fill    = "Hrs",
         boxplot = TRUE)
```

Filter samples on the fly with `filter_eval`:

```{r exps_2}
ExpsPlot(fit_Hotgenes,
         xVar        = "Hrs",
         yVar        = c("CSF2", "IL6"),
         fill        = "Hrs",
         boxplot     = TRUE,
         filter_eval = Hrs != 2)
```

Reorder factor levels using `named_levels`:

```{r exps_3}
ExpsPlot(fit_Hotgenes,
         xVar         = "Hrs",
         yVar         = c("CSF2", "IL6"),
         boxplot      = TRUE,
         fill         = "Hrs",
         named_levels = list(Feature = "IL6",
                             Hrs     = c("6", "2", "0")))
```

---

## 6. `BoxPlot()` — Sample-level QC Plot

`BoxPlot()` renders a boxplot of expression values for each sample.
It is most useful for QC: checking normalization and identifying
outlier samples.

```{r boxplot_1}
BoxPlot(fit_Hotgenes)
```

Restrict to a subset of samples:

```{r boxplot_2}
BoxPlot(fit_Hotgenes,
        SampleIDs = SampleIDs_(fit_Hotgenes)[1:6])
```

---

## 7. `FactoWrapper()` — PCA and Hierarchical Clustering

`FactoWrapper()` runs a full PCA via FactoMineR on the top features
for a given contrast, then clusters samples using HCPC (Hierarchical
Clustering on Principal Components).

```{r pca_run, echo=TRUE, results='hide'}
FactoOutput <- FactoWrapper(
  fit_Hotgenes,
  contrasts   = "Hrs_6_vs_0",
  coldata_ids = c("Hrs", "sh"),
  biplot      = FALSE
)
```

```{r pca_plot, message=FALSE, warning=FALSE}
FactoOutput$res_PPI_pa_1
```

Inspect cluster assignments and top contributing features:

```{r pca_top}
FactoOutput$TopTibble   # top features per cluster
FactoOutput$TopGroups   # cluster membership per sample
```

---

## 8. `coldata_palettes()` — Consistent Colour Schemes

`coldata_palettes()` generates a named list of colour vectors for each
factor in the coldata. This can be passed directly to `DEphe()` or used
in custom ggplot2 themes.

```{r palettes}
coldata_palettes(fit_Hotgenes)
```

---

## Summary of Visualization Functions

| Function | Purpose |
|---|---|
| `DEPlot()` | Bar chart of DE counts across all contrasts |
| `VPlot()` | Volcano plot for a single contrast |
| `Venn_Report()` | Venn diagram of overlapping features |
| `DEphe()` | Heatmap of top hits for a contrast |
| `ExpsPlot()` | Expression trajectory plots |
| `BoxPlot()` | Sample-level expression boxplots (QC) |
| `FactoWrapper()` | PCA + HCPC clustering |
| `coldata_palettes()` | Colour palettes for metadata factors |
