---
title: "Sample-wise Pathway Activity with HotgeneSets"
output:
  github_document:
    html_preview: false
vignette: >
  %\VignetteIndexEntry{Sample-wise Pathway Activity with HotgeneSets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

`HotgeneSets()` computes pathway-level activity scores for each sample
(using GSVA/ssGSEA-family methods) and returns a new Hotgenes object.
This keeps pathway activity analysis separate from the Shiny workflow and
makes it easy to reuse downstream plotting and DE utilities on pathway
features.

```{r set, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  fig.path  = "05_Samplewise_Pathway_Activity_HotgeneSets_files/figure-",
  fig.height = 4, fig.width = 6
)
```

```{r load, message=FALSE}
library(Hotgenes)

fit_Hotgenes <- readRDS(
  system.file("extdata", "fit_Hotgenes.RDS",
              package = "Hotgenes",
              mustWork = TRUE)
)
```

---

## 1. Retrieve Gene Sets

```{r genesets, message=FALSE, warning=FALSE}
H_paths <- msigdbr_wrapper(
  species  = "human",
  set      = c("CP:KEGG"),
  gene_col = "gene_symbol"
)

length(H_paths)
names(H_paths)[1:5]
```

---

## 2. Compute Sample-wise Pathway Activity

```{r hotgenesets, message=FALSE, warning=FALSE}
HotgeneSets_out <- HotgeneSets(
  Hotgenes = fit_Hotgenes,
  geneSets = H_paths,
  kcdf     = "Gaussian",
  method   = "ssgsea",
  minSize  = 2,
  maxSize  = Inf
)

HotgeneSets_out
```

The resulting object stores pathway scores in the expression slot and
supports the same downstream functions used for gene-level analyses.

---

## 3. Explore Pathway-level Differential Activity

```{r pathway_deplot}
DEPlot(HotgeneSets_out, .log2FoldChange = 0, padj_cut = 0.1)
```

```{r pathway_heatmap, fig.height=8, fig.width=8}
DEphe(
  HotgeneSets_out,
  contrasts         = "sh_EWS_vs_Ctrl",
  Topn              = 5,
  annotations       = c("Hrs", "sh"),
  annotation_colors = coldata_palettes(HotgeneSets_out)
)
```

---

## 4. Retrieve Leading Pathway Names for Downstream Use

```{r pathway_names}
pathway_hits <- HotgeneSets_out |>
  DE(
    Report    = "Features",
    contrasts = "sh_EWS_vs_Ctrl",
    padj_cut  = 0.2
  )

head(pathway_hits$sh_EWS_vs_Ctrl)
```

These pathway names can be used directly in additional plots,
reporting, or to initialize a focused Shiny session if needed.
