---
title: "DILI Discovery Proteomics with Hotgenes"
output: 
  github_document:
    html_preview: false

vignette: >
  %\VignetteIndexEntry{DILI Discovery Proteomics with Hotgenes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Building a Hotgenes Object from DILI Discovery Proteomics Data

This vignette demonstrates how to build a Hotgenes object from the Drug-Induced
Liver Injury (DILI) discovery proteomics dataset published in:

> Federspiel et al. (2023). *Proteomics of drug-induced liver injury (DILI)
> identifies a distinct bile acid signature in patients with Hy's Law.*
> **Journal of Hepatology**, 79(2), 370–382.
> <https://doi.org/10.1016/j.jhep.2023.03.013>

The raw data are publicly deposited in MassIVE under accession **MSV000089782**.
A pre-parsed snapshot is shipped with this package as `inst/extdata/dili_raw.RDS`
to avoid a runtime download.

### A note on private metadata

The original analysis used two variables from the private metadata file
`1219_p1-p5_protein_KEY.xlsx` that are **not** available in the public deposit:

1. **subject** – patient identifier, used with `duplicateCorrelation()` to
   control for repeated measures (paired DO/DF samples from the same patient).
   Without subject IDs this blocking step is omitted here.

2. **channel** – exact TMT channel label (e.g. `126C`, `127N`) within each
   pool, used with `voomaByGroup()` to compensate for channel-specific
   variance. `Pool` (P1–P5) is used as a proxy because it captures the
   dominant source of TMT batch variance.

All other analytical steps — VSN normalization, `voomaByGroup`, robust `lmFit`,
and a BH-adjusted p-value threshold of 0.1 — reproduce those described in
Federspiel et al.

```{r set, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.height = 4, fig.width = 6
)
```


## Step 1 — Load data

`dili_raw.RDS` is a pre-parsed snapshot of the public MassIVE deposit
(MSV000089782), saved to `inst/extdata` to avoid a runtime download.

```{r load_data}
library(Hotgenes)

raw_df <- readRDS(
  system.file("extdata", "dili_raw.RDS",
              package = "Hotgenes",
              mustWork = TRUE)
)
```

To re-download and refresh `dili_raw.RDS` directly from MassIVE, run the
following block once (set `eval = FALSE` to skip during package build):

```{r refresh_massive, eval=FALSE}
url <- paste0(
  "https://massive.ucsd.edu/ProteoSAFe/DownloadResultFile?",
  "file=f.MSV000089782%2Fupdates%2F2023-02-28_jfederspiel_1bc96582",
  "%2Fother%2FDILI_discovery_data.xlsx&forceDownload=true"
)

tmp <- tempfile(fileext = ".xlsx")
download.file(url, tmp, mode = "wb")

# Requires readxl and janitor packages (install if needed):
# install.packages(c("readxl", "janitor"))
raw_df <- readxl::read_excel(tmp) |>
  janitor::clean_names()

saveRDS(
  raw_df,
  file = file.path(getwd(), "inst", "extdata", "dili_raw.RDS")
)
```


## Step 2 — Identify sample columns

Sample columns follow the pattern `<condition>_p<pool>_<number>` where
condition is one of `do`, `df`, `hv`, `nafld`, `ndo`, or `ndf`.

```{r sample_cols}
sample_cols <- colnames(raw_df)[
  grepl("^(do|df|hv|nafld|ndo|ndf)_", colnames(raw_df))
]

length(sample_cols)
head(sample_cols)
```


## Step 3 — Filter proteins

Retain only:

* Reviewed UniProt entries (accession starts with `sp|`)
* Proteins that are not contaminants
* Proteins quantified with at least one peptide (`min_peps > 0`)

Unique gene symbols are generated as the `Feature` column, which is required
downstream for GSEA.

```{r filter_proteins}
filtered_exps <- raw_df |>
  dplyr::filter(!grepl("contaminant", .data$protein)) |>
  dplyr::filter(grepl("^sp[|]", .data$protein)) |>
  dplyr::filter(.data$min_peps > 0) |>
  dplyr::mutate(Feature = make.names(.data$gene_symbol, unique = TRUE))

nrow(filtered_exps)
```


## Step 4 — Build the expression matrix

```{r expr_matrix}
expr_matrix <- filtered_exps |>
  dplyr::select("Feature", dplyr::any_of(sample_cols)) |>
  tibble::column_to_rownames("Feature") |>
  as.matrix()

dim(expr_matrix)
```


## Step 5 — Build the protein ID mapper

The mapper links `Feature` names back to the original gene symbol, full
protein accession, and description. Hotgenes functions use the mapper to
display human-readable labels in plots and tables.

```{r mapper}
mapper_df <- filtered_exps |>
  dplyr::select(
    "Feature",
    "Gene"        = "gene_symbol",
    "Protein"     = "protein",
    "Description" = "description"
  )

head(mapper_df)
```


## Step 6 — Build sample metadata (coldata)

Condition and pool are parsed from the sample column names.  
`Condition` levels are ordered so that healthy volunteers (`HV`) appear first.

```{r coldata}
coldata <- data.frame(
  Sample    = sample_cols,
  Condition = toupper(sub("_p[0-9]+_[0-9]+$", "", sample_cols)),
  Pool      = toupper(sub(".*_(p[0-9]+)_.*", "\\1", sample_cols)),
  row.names = sample_cols,
  stringsAsFactors = TRUE
)

coldata[["Condition"]] <- factor(
  coldata[["Condition"]],
  levels = c("HV", "DO", "DF", "NDO", "NDF", "NAFLD")
)

table(coldata$Condition)
```


## Step 7 — VSN normalization

Variance-stabilising normalisation (`limma::normalizeVSN`) is applied to
correct for intensity-dependent variance across the dynamic range of the TMT
measurements. Missing values are replaced with 0 before normalisation.

```{r vsn_norm}
expr_matrix[is.na(expr_matrix)] <- 0
expr_matrix_vsn <- limma::normalizeVSN(expr_matrix)
```


## Step 8 — Design matrix

A no-intercept design matrix is constructed so that each coefficient directly
represents the mean VSN-normalised expression for one condition, making
contrasts straightforward to specify.

```{r design}
design <- model.matrix(~ 0 + Condition, data = coldata)
colnames(design) <- gsub("Condition", "", colnames(design))

head(design)
```


## Step 9 — voomaByGroup transformation

`limma::voomaByGroup` estimates mean–variance trends separately for each TMT
pool and converts the VSN-normalised intensities into precision weights for
downstream linear modelling. `Pool` is used as the grouping variable because
it captures the dominant source of batch variance in the public deposit.

```{r vooma}
vm_exp <- limma::voomaByGroup(
  y      = expr_matrix_vsn,
  group  = coldata[["Pool"]],
  design = design,
  plot   = FALSE
)
```


## Step 10 — Robust linear model fit

`limma::lmFit` with `method = "robust"` downweights outlier samples when
estimating per-protein coefficients, improving robustness against occasional
poorly quantified channels.

Note: `duplicateCorrelation()` is omitted here because the patient subject IDs
required to model repeated measures (paired DO/DF samples) are not available
in the public deposit.

```{r lmfit}
fit <- limma::lmFit(
  vm_exp,
  design,
  method = "robust"
)
```


## Step 11 — Define contrasts and run eBayes

Seven clinically motivated contrasts are tested, all relative to healthy
volunteers (`HV`), plus one intra-DILI comparison (`DF_vs_DO`) and one
cross-group comparison (`NDO_vs_DO`).

```{r contrasts}
contrasts_mat <- limma::makeContrasts(
  DO_vs_HV    = DO    - HV,
  DF_vs_HV    = DF    - HV,
  NDO_vs_HV   = NDO   - HV,
  NDF_vs_HV   = NDF   - HV,
  NAFLD_vs_HV = NAFLD - HV,
  DF_vs_DO    = DF    - DO,
  NDO_vs_DO   = NDO   - DO,
  levels = design
)

fit2 <- limma::contrasts.fit(fit, contrasts_mat)
fit2 <- limma::eBayes(fit2)
```


## Step 12 — Create the Hotgenes object

All upstream objects are bundled into a single `Hotgenes` object using
`Hotgeneslimma()`. The raw log2-transformed matrix is stored as an auxiliary
assay so it is available for alternative visualisations.

```{r hotgenes_object}
dili_hotgenes <- Hotgeneslimma(
  limmafit        = fit2,
  coldata         = coldata,
  Expression      = vm_exp,
  Expression_name = "VSN",
  Exps_list       = list(log2 = log2(expr_matrix + 1)),
  Mapper          = mapper_df
)

dili_hotgenes
```


## What can you do with the DILI Hotgenes object?

### Summary of differentially expressed proteins

```{r de_summary}
DE(dili_hotgenes, Topn = 5)
```

### Overview plot of all contrasts

```{r dep_plot}
DEPlot(dili_hotgenes, .log2FoldChange = 0, padj_cut = 0.1)
```

### Volcano plot

```{r volcano}
VPlot(dili_hotgenes,
      .log2FoldChange = 1, padj_cut = 0.1,
      contrasts = "DO_vs_HV")
```

### Venn diagram of overlapping hits

```{r venn, fig.height=8, fig.width=8}
dili_hotgenes |>
  DE(
    Report     = "Features",
    contrasts  = c("DO_vs_HV", "DF_vs_HV", "NAFLD_vs_HV"),
    .log2FoldChange = 0, padj_cut = 0.1
  ) |>
  Venn_Report()
```

### Heatmap of top hits

```{r heatmap, fig.height=8, fig.width=8}
DEphe(dili_hotgenes,
      contrasts  = "DO_vs_HV",
      Topn       = 5,
      cellheight = 10,
      cellwidth  = 8,
      annotation_colors = coldata_palettes(dili_hotgenes),
      annotations = c("Condition", "Pool"))
```

### Expression plots

```{r exps_plot}
ExpsPlot(dili_hotgenes,
  xVar    = "Condition",
  yVar    = c("APOB", "ALB"),
  fill    = "Condition",
  boxplot = TRUE
)
```

### GSEA with KEGG pathways

```{r gsea}
# Get KEGG gene sets
H_paths <- msigdbr_wrapper(
  species  = "human",
  set      = "CP:KEGG_MEDICUS",
  gene_col = "gene_symbol"
)

# Rank genes for the DO_vs_HV contrast
InputRanks <- dili_hotgenes |>
  DE(
    Report    = "Ranks",
    contrasts = "DO_vs_HV",
    Rank_name = "Gene",
    padj_cut  = 0.1
  )

# Run fgsea
Out_GSEA <- fgsea_(
  Ranks    = InputRanks,
  pathways = H_paths,
  nproc    = 1,
  minSize  = 5,
  maxSize  = Inf
)

# Summarise significant pathways
Out_GSEA |>
  fgsea_Results(
    contrasts = "DO_vs_HV",
    padj_cut  = 0.2,
    mode      = "D"
  )
```

### Interactive exploration with Shiny

```{r shiny, eval=FALSE}
Shiny_Hotgenes(dili_hotgenes)
```
